### 2018-05-22 paper.tex
Début du papier: abstract, intro, références + Makefile

\documentclass[twocolumn]{article}
\usepackage[utf8]{inputenc}

\author{
	K\'evin Le Bon
	\and
	Alan Schmitt
}

\title{MLExplain}

\begin{document}
\maketitle

\begin{abstract}
	MLExplain is a step-by-step visual interpreter for OCaml that enables the user
	to inspect both their program's state and the interpreter's state itself.
\end{abstract}

\section{Introduction}

The semantics of a programming language can be very complex. When a language has no specification,
the semantics is then defined by the implementations of the language, i.e. interpreters and
compilers. However even when a specification is available, it can be difficult to understand why
the execution of a specific program results to a certain output.

The project \emph{JSExplain}\footnote{https://github.com/jscert/jsexplain} aims to describe
a JavaScript program's execution by showing every step of an interpreter which behavior is the
closest to JavaScript specification. This paper shows how we adapted JSExplain to OCaml.

Unlike JavaScript, OCaml has no official specification. However, OCaml code execution is
fairly simple because the amount of language constructions is small. We have written an
interpreter for OCaml's typed abstract syntax tree (AST). This AST is close to source code and
need little transformations but supplies us with resolved names -- which is useful for module and
signatures inclusion and absolutely mandatory for features like named parameters in functions.
The semantic we give to OCaml is higher level than the one described in \emph{ZINC}
\cite{Leroy-ZINC} virtual machine which is at the same level as object code.

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2018-05-22 paper.tex
Écrit une bonne partie de la section sur JSExplain

@@
+%\documentclass{article}
 \documentclass[twocolumn]{article}
@@
 	MLExplain is a step-by-step visual interpreter for OCaml that enables the user
-	to inspect both their program's state and the interpreter's state itself.
+	to inspect both their program's state and the interpreter's state itself. This interpreter
+	aims to provide the user with a better understanding of OCaml's semantics.
 \end{abstract}
@@
 
+\section{The visual interpreter JSExplain}
+
+An interpreter is a program that takes code written in a specific language as input
+and execute it. In order to achieve its goal, the interpreter does several analyses ;
+here are the most common : lexical analysis, parsing and typing.
+
+We call a \emph{visual interpreter} an interpreter that is able to display the program's state,
+the current location in source and local-variable values at each step of the execution as well as
+the state of the interpreter itself. JSExplain is a visual interpreter for JavaScript. The aim of
+this tool is to enable a user to better understand JavaScript's semantics. JSExplain's interface
+highlights the expression of the program currently evaluated as well as the instruction of
+the interpreter that is executed.
+
+\section{JSExplain functioning}
+
+JSExplain is a visual interpreter. To display its state, the interpreter is compiled with a
+compiler that adds calls to a tracing function. The traced retrieved during the execution of the
+program gives many information about the state of the interpreter:
+
+\begin{itemize}
+	\item existing variables and their values
+	\item return values of functions of the interpreter
+\end{itemize}
+
+JSExplain's web interface uses these traces in order to display the program and interpreter state
+at every step of the execution.
+
+\subsection{JSExplain's compiler}
+
+% Expliquer que JSExplain est écrit en OCaml et compilé vers JavaScript pour être intégré dans
+% la page web qui va avec
+
+JSExplain needs its own OCaml to JavaScript compiler in order to place the calls to
+\verb|log_event| which is the tracing function. This function is also used to get the location
+of the different syntactical objects in the source file in order to produce code highlighting in
+JSExplain's interface.
+
 \bibliographystyle{plain}

### 2018-05-23 paper.tex
Refactor pour faire tenir le code sur 80 colonnes

@@
 
-\author{
-	K\'evin Le Bon
-	\and
-	Alan Schmitt
-}
+\author{ K\'evin Le Bon \and Alan Schmitt }
 
@@
 \begin{abstract}
-	MLExplain is a step-by-step visual interpreter for OCaml that enables the user
-	to inspect both their program's state and the interpreter's state itself. This interpreter
-	aims to provide the user with a better understanding of OCaml's semantics.
+  MLExplain is a step-by-step visual interpreter for OCaml that enables the
+  user to inspect both their program's state and the interpreter's state
+  itself. This interpreter aims to provide the user with a better understanding
+  of OCaml's semantics.
 \end{abstract}
@@
 
-The semantics of a programming language can be very complex. When a language has no specification,
-the semantics is then defined by the implementations of the language, i.e. interpreters and
-compilers. However even when a specification is available, it can be difficult to understand why
-the execution of a specific program results to a certain output.
+The semantics of a programming language can be very complex. When a language
+has no specification, the semantics is then defined by the implementations of
+the language, i.e. interpreters and compilers. However even when a
+specification is available, it can be difficult to understand why the execution
+of a specific program results to a certain output.
 
-The project \emph{JSExplain}\footnote{https://github.com/jscert/jsexplain} aims to describe
-a JavaScript program's execution by showing every step of an interpreter which behavior is the
-closest to JavaScript specification. This paper shows how we adapted JSExplain to OCaml.
+The project \emph{JSExplain}\footnote{https://github.com/jscert/jsexplain} aims
+to describe a JavaScript program's execution by showing every step of an
+interpreter which behavior is the closest to JavaScript specification. This
+paper shows how we adapted JSExplain to OCaml.
 
-Unlike JavaScript, OCaml has no official specification. However, OCaml code execution is
-fairly simple because the amount of language constructions is small. We have written an
-interpreter for OCaml's typed abstract syntax tree (AST). This AST is close to source code and
-need little transformations but supplies us with resolved names -- which is useful for module and
-signatures inclusion and absolutely mandatory for features like named parameters in functions.
-The semantic we give to OCaml is higher level than the one described in \emph{ZINC}
-\cite{Leroy-ZINC} virtual machine which is at the same level as object code.
+Unlike JavaScript, OCaml has no official specification. However, OCaml code
+execution is fairly simple because the amount of language constructions is
+small. We have written an interpreter for OCaml's typed abstract syntax tree
+(AST). This AST is close to source code and need little transformations but
+supplies us with resolved names -- which is useful for module and signatures
+inclusion and absolutely mandatory for features like named parameters in
+functions.  The semantic we give to OCaml is higher level than the one
+described in \emph{ZINC} \cite{Leroy-ZINC} virtual machine which is at the same
+level as object code.
 
@@
 
-An interpreter is a program that takes code written in a specific language as input
-and execute it. In order to achieve its goal, the interpreter does several analyses ;
-here are the most common : lexical analysis, parsing and typing.
+An interpreter is a program that takes code written in a specific language as
+input and execute it. In order to achieve its goal, the interpreter does
+several analyses ; here are the most common : lexical analysis, parsing and
+typing.
 
-We call a \emph{visual interpreter} an interpreter that is able to display the program's state,
-the current location in source and local-variable values at each step of the execution as well as
-the state of the interpreter itself. JSExplain is a visual interpreter for JavaScript. The aim of
-this tool is to enable a user to better understand JavaScript's semantics. JSExplain's interface
-highlights the expression of the program currently evaluated as well as the instruction of
-the interpreter that is executed.
+We call a \emph{visual interpreter} an interpreter that is able to display the
+program's state, the current location in source and local-variable values at
+each step of the execution as well as the state of the interpreter itself.
+JSExplain is a visual interpreter for JavaScript. The aim of this tool is to
+enable a user to better understand JavaScript's semantics. JSExplain's
+interface highlights the expression of the program currently evaluated as well
+as the instruction of the interpreter that is executed.
 
@@
 
-JSExplain is a visual interpreter. To display its state, the interpreter is compiled with a
-compiler that adds calls to a tracing function. The traced retrieved during the execution of the
-program gives many information about the state of the interpreter:
+JSExplain is a visual interpreter. To display its state, the interpreter is
+compiled with a compiler that adds calls to a tracing function. The traced
+retrieved during the execution of the program gives many information about the
+state of the interpreter:
 
@@
 
-JSExplain's web interface uses these traces in order to display the program and interpreter state
-at every step of the execution.
+JSExplain's web interface uses these traces in order to display the program and
+interpreter state at every step of the execution.
 
@@
 
-% Expliquer que JSExplain est écrit en OCaml et compilé vers JavaScript pour être intégré dans
-% la page web qui va avec
+% Référence pour coq ?
+JSExplain was first written in Coq. OCaml code has been extracted from the Coq
+source and is now the primary source code. The interpreter is compiled into
+JavaScript in order to use it within a web page. For this reason, the project
+embeds a homemade OCaml-to-JavaScript compiler.
 
-JSExplain needs its own OCaml to JavaScript compiler in order to place the calls to
-\verb|log_event| which is the tracing function. This function is also used to get the location
-of the different syntactical objects in the source file in order to produce code highlighting in
-JSExplain's interface.
+JSExplain also needs its own OCaml to JavaScript compiler in order to place the
+calls to \verb|log_event| which is the tracing function. This function is also
+used to get the location of the different syntactical objects in the source
+file in order to produce code highlighting in JSExplain's interface.
 

### 2018-05-23 paper.tex
Small tweaks
jsexplain bib entry
mlexplain architecture

--- 2018-05-22 paper.tex
@@
 \usepackage[utf8]{inputenc}
+\usepackage{tikz}
+\usetikzlibrary{positioning,fit,shapes,arrows.meta}
 
@@
 	to inspect both their program's state and the interpreter's state itself. This interpreter
-	aims to provide the user with a better understanding of OCaml's semantics.
+	aims to provide the user with a better understanding of the semantics of OCaml.
 \end{abstract}
@@
 The semantics of a programming language can be very complex. When a language has no specification,
-the semantics is then defined by the implementations of the language, i.e. interpreters and
+the semantics is then defined by the implementations of the language, i.e., interpreters and
 compilers. However even when a specification is available, it can be difficult to understand why
@@
 
-The project \emph{JSExplain}\footnote{https://github.com/jscert/jsexplain} aims to describe
+The project \emph{JSExplain} \cite{chargueraud:hal-01745792} aims to describe
 a JavaScript program's execution by showing every step of an interpreter which behavior is the
@@
 
+\begin{figure*}
+  \centering
+  \begin{tikzpicture}[nodes = {align = center}]
+    \node (focaml) [draw, text width=2cm] {Interpreter and libraries (OCaml)};
+
+    \node (fjs) [draw, right = 1cm of focaml, text width=2cm, rectangle split,
+    rectangle split parts=2, text centered] {Libraries (JS) \nodepart{second}
+      Interpreter with traces (JS)};
+
+    \node (ast) [draw, above = of fjs, text width=2cm] {AST of interpreted
+      program};
+
+    \node (source) [draw, right = 2cm of ast, text width=2cm, text centered]
+    {interpreted program};
+
+    \node [draw, dotted, fit=(fjs) (ast) (source)] {web page};
+
+    \node (trace) [draw, right = 2cm of fjs, text width=2cm, text centered]
+    {trace};
+
+    \path[->,thick,>=stealth] (focaml.north
+    east) edge[bend left] node[above left] {compiler} (fjs.north west) ;
+
+    \path[->,thick,>=stealth] (focaml.south east) edge[bend right] node[below
+    left] {\parbox{2cm}{\centering tracing compiler}} (fjs.south west);
+
+    \path[->,thick,dotted,>=stealth] (source) edge node [above] {Parser} node [below] {Js\_of\_ocaml} (ast);
+
+    \path[->,thick,dotted,>=stealth] (ast) edge (fjs);
+
+    \path[->,thick,dotted,>=stealth] (ast) edge (fjs);
+
+    \path[->,thick,dotted,>=stealth] (fjs) edge (trace);
+  \end{tikzpicture}
+
+  \caption{Architecture of MLExplain}
+  \label{fig:archi}
+\end{figure*}
+
 \bibliographystyle{plain}

### 2018-05-23 paper.tex

@@
 \begin{abstract}
-	MLExplain is a step-by-step visual interpreter for OCaml that enables the user
-	to inspect both their program's state and the interpreter's state itself. This interpreter
-	aims to provide the user with a better understanding of the semantics of OCaml.
+  MLExplain is a step-by-step visual interpreter for OCaml that enables the
+  user to inspect both their program's state and the interpreter's state
+  itself. This interpreter aims to provide the user with a better understanding
+  of the semantics of OCaml semantics.
 \end{abstract}
@@
 
-The semantics of a programming language can be very complex. When a language has no specification,
-the semantics is then defined by the implementations of the language, i.e., interpreters and
-compilers. However even when a specification is available, it can be difficult to understand why
-the execution of a specific program results to a certain output.
+The semantics of a programming language can be very complex. When a language
+has no specification, the semantics is then defined by the implementations of
+the language, i.e., interpreters and compilers. However even when a
+specification is available, it can be difficult to understand why the execution
+of a specific program results to a certain output.
 
-The project \emph{JSExplain} \cite{chargueraud:hal-01745792} aims to describe
-a JavaScript program's execution by showing every step of an interpreter which behavior is the
-closest to JavaScript specification. This paper shows how we adapted JSExplain to OCaml.
+The project \emph{JSExplain} \cite{chargueraud:hal-01745792} aims
+to describe a JavaScript program's execution by showing every step of an
+interpreter which behavior is the closest to JavaScript specification. This
+paper shows how we adapted JSExplain to OCaml.
 
-Unlike JavaScript, OCaml has no official specification. However, OCaml code execution is
-fairly simple because the amount of language constructions is small. We have written an
-interpreter for OCaml's typed abstract syntax tree (AST). This AST is close to source code and
-need little transformations but supplies us with resolved names -- which is useful for module and
-signatures inclusion and absolutely mandatory for features like named parameters in functions.
-The semantic we give to OCaml is higher level than the one described in \emph{ZINC}
-\cite{Leroy-ZINC} virtual machine which is at the same level as object code.
+Unlike JavaScript, OCaml has no official specification. However, OCaml code
+execution is fairly simple because the amount of language constructions is
+small. We have written an interpreter for OCaml's typed abstract syntax tree
+(AST). This AST is close to source code and need little transformations but
+supplies us with resolved names -- which is useful for module and signatures
+inclusion and absolutely mandatory for features like named parameters in
+functions.  The semantic we give to OCaml is higher level than the one
+described in \emph{ZINC} \cite{Leroy-ZINC} virtual machine which is at the same
+level as object code.
 

### 2018-05-23 paper.tex
Continue d'écrire sur JSExplain

--- 2018-05-23 paper.tex
@@
 
+The whole project takes the shape of a web application composed of the
+JavaScript interpreter and a view. The view allows the user to write JavaScript
+code and run it. It also displays the state of the program and the state of the
+interpreter.
+
+The program's execution is not interactive: when the button \texttt{run} is
+pressed, the program is executed all at one and the traces are generated. Only
+then, the user can replay the execution step by step.
+
 \subsection{JSExplain's compiler}
@@
 
+This compiler offers several compilation modes:
+
+\begin{itemize}
+  \item \texttt{unlog}: generate JavaScript without tracing.
+  \item \texttt{log}: generate JavaScript with tracing.
+  \item \texttt{mlloc}, \texttt{token} and \texttt{ptoken}: generate files with
+    informations about syntastical elements from the input file.
+  \item \texttt{pseudo}: generate JavaScript enhanced with syntactic sugar for
+    pattern matching or monadic operations -- it is supposed to be easier to
+    read.
+\end{itemize}
+
+\subsubsection{fOCaml}
+This compiler does not recognized the whole OCaml language, it can only handle
+a small subset we named \texttt{fOCaml}. Among other restrictions, it cannot
+handle objects and imperative features, exceptions, inner modules, and tuples.
+
 \bibliographystyle{plain}

### 2018-05-24 paper.tex

--- 2018-05-23 paper.tex
@@
 
-\subsection{JSExplain's compiler}
+% J'aime pas trop la voir tout en bas, mais ici ça fait bizarre, si tu as mieux
+% je prends.
 
-% Référence pour coq ?
-JSExplain was first written in Coq. OCaml code has been extracted from the Coq
-source and is now the primary source code. The interpreter is compiled into
-JavaScript in order to use it within a web page. For this reason, the project
-embeds a homemade OCaml-to-JavaScript compiler.
-
-JSExplain also needs its own OCaml to JavaScript compiler in order to place the
-calls to \verb|log_event| which is the tracing function. This function is also
-used to get the location of the different syntactical objects in the source
-file in order to produce code highlighting in JSExplain's interface.
-
-\begin{figure*}
+\begin{figure*}[ht]
   \centering
@@
 
-    \path[->,thick,dotted,>=stealth] (source) edge node [above] {Parser} node [below] {Js\_of\_ocaml} (ast);
+    \path[->,thick,dotted,>=stealth] (source) edge node [above] {Parser} node
+    [below] {Js\_of\_ocaml} (ast);
 
@@
 
+The whole project takes the shape of a web application composed of the
+JavaScript interpreter and a view. The view allows the user to write JavaScript
+code and run it. It also displays the state of the program and the state of the
+interpreter.
+
+The program's execution is not interactive: when the button \texttt{run} is
+pressed, the program is executed all at one and the traces are generated. Only
+then, the user can replay the execution step by step.
+
+\subsection{JSExplain's compiler}
+
+% Référence pour coq ?
+JSExplain was first written in Coq. OCaml code has been extracted from the Coq
+source and is now the primary source code. The interpreter is compiled into
+JavaScript in order to use it within a web page. For this reason, the project
+embeds a homemade OCaml-to-JavaScript compiler.
+
+JSExplain also needs its own OCaml to JavaScript compiler in order to place the
+calls to \verb|log_event| which is the tracing function. This function is also
+used to get the location of the different syntactical objects in the source
+file in order to produce code highlighting in JSExplain's interface.
+
+This compiler offers several compilation modes:
+
+\begin{itemize}
+  \item \texttt{unlog}: generate JavaScript without tracing.
+  \item \texttt{log}: generate JavaScript with tracing.
+  \item \texttt{mlloc}, \texttt{token} and \texttt{ptoken}: generate files with
+    informations about syntastical elements from the input file.
+  \item \texttt{pseudo}: generate JavaScript enhanced with syntactic sugar for
+    pattern matching or monadic operations -- it is supposed to be easier to
+    read.
+\end{itemize}
+
+\subsubsection{fOCaml}
+This compiler does not recognized the whole OCaml language, it can only handle
+a small subset we named \texttt{fOCaml}. Among other restrictions, it cannot
+handle objects and imperative features, exceptions, inner modules, and tuples.
+
 \bibliographystyle{plain}

### 2018-05-24 paper.tex
Rebougé la figure et continué sur MLExplain
j'avais pas compris que la figure présentait MLExplain et non JSExplain.

@@
 
-% J'aime pas trop la voir tout en bas, mais ici ça fait bizarre, si tu as mieux
-% je prends.
+The whole project takes the shape of a web application composed of the
+JavaScript interpreter and a view. The view allows the user to write JavaScript
+code and run it. It also displays the state of the program and the state of the
+interpreter.
+
+The program's execution is not interactive: when the button \texttt{run} is
+pressed, the program is executed all at one and the traces are generated. Only
+then, the user can replay the execution step by step.
+
+\subsection{JSExplain's compiler}
+
+% Référence pour coq ?
+JSExplain was first written in Coq. OCaml code has been extracted from the Coq
+source and is now the primary source code. The interpreter is compiled into
+JavaScript in order to use it within a web page. For this reason, the project
+embeds a homemade OCaml-to-JavaScript compiler.
+
+JSExplain also needs its own OCaml to JavaScript compiler in order to place the
+calls to \verb|log_event| which is the tracing function. This function is also
+used to get the location of the different syntactical objects in the source
+file in order to produce code highlighting in JSExplain's interface.
+
+This compiler offers several compilation modes:
+
+\begin{itemize}
+  \item \texttt{unlog}: generate JavaScript without tracing.
+  \item \texttt{log}: generate JavaScript with tracing.
+  \item \texttt{mlloc}, \texttt{token} and \texttt{ptoken}: generate files with
+    informations about syntastical elements from the input file.
+  \item \texttt{pseudo}: generate JavaScript enhanced with syntactic sugar for
+    pattern matching or monadic operations -- it is supposed to be easier to
+    read.
+\end{itemize}
+
+Modes \texttt{unlog} and \texttt{log} exist so that auxiliary code in the
+interpreter -- like linked-list or hash-map internal-management functions -- is
+not traced.  The user is thus able to concentrate on the logical part of the
+interpreter functioning and forget about implementation details.
+
+\subsubsection{fOCaml}
+This compiler does not recognized the whole OCaml language, it can only handle
+a small subset we named \texttt{fOCaml}. Among other restrictions, it cannot
+handle objects and imperative features, exceptions, inner modules, and tuples.
+The main reason for this limitation is that compiling OCaml into JavaScript is
+not trivial, many features -- e.g. nested pattern matching -- are hard to
+represent to JavaScript.
+
+The fOCaml code is compiled into \texttt{fJs}, a functional subset of
+JavaScript. The compiler preserves as much as possible the code structure to
+enable the user to understand both fOCaml and fJs codes as the same one.
+
+% Petite explication sur le fait que du coup on écrit du code fonctionnel pur
+% et les implications que ça a ?
+
+\section{MLExplain}
+
+At first glance, JSExplain seems tightly linked to JavaScript. However, the
+architecture of the project actually allows us to take off the interpreter
+completely and replace it with another one.
+
+% Plus de blabla
+
+\subsection{Two compilers}
+
+To interpret OCaml code, we need to parse it. However, our compiler does not
+allow us to use the standard library nor an external one.
 
@@
 
-The whole project takes the shape of a web application composed of the
-JavaScript interpreter and a view. The view allows the user to write JavaScript
-code and run it. It also displays the state of the program and the state of the
-interpreter.
-
-The program's execution is not interactive: when the button \texttt{run} is
-pressed, the program is executed all at one and the traces are generated. Only
-then, the user can replay the execution step by step.
-
-\subsection{JSExplain's compiler}
-
-% Référence pour coq ?
-JSExplain was first written in Coq. OCaml code has been extracted from the Coq
-source and is now the primary source code. The interpreter is compiled into
-JavaScript in order to use it within a web page. For this reason, the project
-embeds a homemade OCaml-to-JavaScript compiler.
-
-JSExplain also needs its own OCaml to JavaScript compiler in order to place the
-calls to \verb|log_event| which is the tracing function. This function is also
-used to get the location of the different syntactical objects in the source
-file in order to produce code highlighting in JSExplain's interface.
-
-This compiler offers several compilation modes:
-
-\begin{itemize}
-  \item \texttt{unlog}: generate JavaScript without tracing.
-  \item \texttt{log}: generate JavaScript with tracing.
-  \item \texttt{mlloc}, \texttt{token} and \texttt{ptoken}: generate files with
-    informations about syntastical elements from the input file.
-  \item \texttt{pseudo}: generate JavaScript enhanced with syntactic sugar for
-    pattern matching or monadic operations -- it is supposed to be easier to
-    read.
-\end{itemize}
-
-\subsubsection{fOCaml}
-This compiler does not recognized the whole OCaml language, it can only handle
-a small subset we named \texttt{fOCaml}. Among other restrictions, it cannot
-handle objects and imperative features, exceptions, inner modules, and tuples.
-
 \bibliographystyle{plain}

### 2018-05-24 paper.tex
Rajouté une référence pour js_of_ocaml

@@
 To interpret OCaml code, we need to parse it. However, our compiler does not
-allow us to use the standard library nor an external one.
+allow us to use the standard library nor an external one. Fortunately,
+JSExplain itself did not trace the parsing process. That is why we have been
+able to separate totally the parser from the interpreter and use another
+compiler to compile it. The figure \ref{fig:archi} describes the architecture
+of the whole application.
+
+The parser is compiled with
+\emph{js\_of\_ocaml} \cite{DBLP:journals/spe/VouillonB14}, the compiler of the
+project \emph{Ocsigen}.
 

### 2018-05-24 paper.tex
Rajouté une référence pour Ocsigen

--- 2018-05-24 paper.tex
@@
 
-To interpret OCaml code, we need to parse it. However, our compiler does not
-allow us to use the standard library nor an external one.
+To interpret OCaml code, we need to parse and type it. However, our compiler
+does not allow us to use the standard library nor an external one. Fortunately,
+JSExplain itself did not trace both parsing and typing processes. That is why
+we have been able to separate totally the frontend from the interpreter itself
+and use another compiler to compile it. The figure \ref{fig:archi} describes
+the architecture of the whole application.
+
+The frontend is compiled with \emph{js\_of\_ocaml}
+\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
+\emph{Ocsigen} \cite{balat:hal-00691710}.
 

### 2018-05-24 paper.tex
adding details to JSExplain, WIP

--- 2018-05-24 paper.tex
@@
 \begin{abstract}
-  MLExplain is a step-by-step visual interpreter for OCaml that enables the
+  MLExplain is a step-by-step interpreter for OCaml that enables the
   user to inspect both their program's state and the interpreter's state
@@
 
-The project \emph{JSExplain} \cite{chargueraud:hal-01745792} aims
-to describe a JavaScript program's execution by showing every step of an
-interpreter which behavior is the closest to JavaScript specification. This
-paper shows how we adapted JSExplain to OCaml.
+The goal of the \emph{JSExplain} project \cite{chargueraud:hal-01745792} is to
+describe the execution of a JavaScript program by showing every step of an
+interpreter whose behavior is very closest to the specification of JavaScript.
+In this paper, we show how we adapted JSExplain to OCaml.
 
@@
 small. We have written an interpreter for OCaml's typed abstract syntax tree
-(AST). This AST is close to source code and need little transformations but
+(AST). This AST is close to the source code and need little transformations but
 supplies us with resolved names -- which is useful for module and signatures
-inclusion and absolutely mandatory for features like named parameters in
-functions.  The semantic we give to OCaml is higher level than the one
-described in \emph{ZINC} \cite{Leroy-ZINC} virtual machine which is at the same
-level as object code.
+inclusion and absolutely mandatory for features like named parameters and
+optional parameters in functions. The semantic we give to OCaml is higher level
+than the one described in \emph{ZINC} \cite{Leroy-ZINC} virtual machine which is
+at the same level as bytecode.
 
-\section{The visual interpreter JSExplain}
+\section{JSExplain}
 
-An interpreter is a program that takes code written in a specific language as
-input and execute it. In order to achieve its goal, the interpreter does
-several analyses ; here are the most common : lexical analysis, parsing and
-typing.
+\subsection{Double Debugger}
+\label{subsec:double-debugger}
 
-We call a \emph{visual interpreter} an interpreter that is able to display the
-program's state, the current location in source and local-variable values at
-each step of the execution as well as the state of the interpreter itself.
-JSExplain is a visual interpreter for JavaScript. The aim of this tool is to
-enable a user to better understand JavaScript's semantics. JSExplain's
-interface highlights the expression of the program currently evaluated as well
-as the instruction of the interpreter that is executed.
+We call a \emph{double debugger} an interpreter that is able to run step by step
+and display the program's state and current location in source, as well as the
+state and location in the source of the interpreter itself. JSExplain is a
+double debugger for JavaScript. The aim of this tool is to help a user to better
+understand JavaScript's semantics, by providing an interpreter that is very
+close to the specification of JavaScript. JSExplain's interface highlights the
+expression of the program currently evaluated as well as the instruction of the
+interpreter that is executed.
 
-\section{JSExplain functioning}
+\subsection{Architecture}
+
+The core of JSExplain is a JavaScript interpreter written in a subset of OCaml.
+This interpreter is compiled to a subset of JavaScript and instrumented to
+generate a trace of its execution when run. This trace can then be navigated
+using a web-based
+tool.\footnote{https://jscert.github.io/jsexplain/branch/master/driver.html}
+
+The subset of OCaml we support for writing the interpreter is purely functional
+with variables, constants, sequence, conditional, let-binding, function
+definition, function application (with support for prefix and infix functions),
+data constructors, records (including record projections, and the
+``record-with'' construct to build a copy of a record with a number of fields
+updated), tuples (i.e., anonymous records), and simple pattern matching (only
+with non-nested patterns, restricted to data constructors, constants, variables,
+and wildcards). For convenience, let-bindings and functions may bind simple
+patterns (as opposed to only variables). We also support \texttt{ppx} extensions
+for monadic programming, to simplify the handling of errors.
+
+The target language of our compiler is a purely functional subset of JavaScript
+where there is never any type conversion, where objects are used as records (no
+use of their \texttt{prototype} field), and where tuples are encoded into
+arrays.
+
+ALAN: stopped here
 

### 2018-05-24 paper.tex

@@
 
-To interpret OCaml code, we need to parse it. However, our compiler does not
-allow us to use the standard library nor an external one. Fortunately,
-JSExplain itself did not trace the parsing process. That is why we have been
-able to separate totally the parser from the interpreter and use another
-compiler to compile it. The figure \ref{fig:archi} describes the architecture
-of the whole application.
+To interpret OCaml code, we need to parse and type it. However, our compiler
+does not allow us to use the standard library nor an external one. Fortunately,
+JSExplain itself did not trace both parsing and typing processes. That is why
+we have been able to separate totally the frontend from the interpreter itself
+and use another compiler to compile it. The figure \ref{fig:archi} describes
+the architecture of the whole application.
 
-The parser is compiled with
-\emph{js\_of\_ocaml} \cite{DBLP:journals/spe/VouillonB14}, the compiler of the
-project \emph{Ocsigen}.
+The frontend is compiled with \emph{js\_of\_ocaml}
+\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
+\emph{Ocsigen} \cite{balat:hal-00691710}.
 

### 2018-05-24 paper.tex
Ajouté une phrase, commit rapide
Ajouté une petite partie sur l'interface web

@@
 \cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
-\emph{Ocsigen} \cite{balat:hal-00691710}.
+\emph{Ocsigen} \cite{balat:hal-00691710}. The largest part of the frontend is a
+module that serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a
+JavaScript object compatible with the backend interpreter's own representation
+of the AST. We use the \texttt{compiler-libs} library to do the actual parsing
+and typing. We ensure that the typed AST we get is the same as the official
+OCaml compiler AST since we use the exact same frontend.
+
+% Toute une partie sur MLExplain lui-même...
+
+\subsection{Web interface changes}
+
+The data structures manipulated by MLExplain are very different from those
+manipulated by JSExplain since they don't interpret the same language. This led
+us to modify the web interface so that it can recognize the different syntactic
+elements of OCaml and pretty print them correctly as well as OCaml values.
+
+Fortunately the web interface is very simple. It is composed of some auxiliary
+JavaScript files, the main JavaScript file containing the logic of the
+application and the corresponding HTML page.
 

### 2018-05-24 paper.tex
Continué

@@
 
-To interpret OCaml code, we need to parse and type it. However, our compiler
-does not allow us to use the standard library nor an external one. Fortunately,
-JSExplain itself did not trace both parsing and typing processes. That is why
-we have been able to separate totally the frontend from the interpreter itself
-and use another compiler to compile it. The figure \ref{fig:archi} describes
-the architecture of the whole application.
-
-The frontend is compiled with \emph{js\_of\_ocaml}
-\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
-\emph{Ocsigen} \cite{balat:hal-00691710}. The largest part of the frontend is a
-module that serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a
-JavaScript object compatible with the backend interpreter's own representation
-of the AST. We use the \texttt{compiler-libs} library to do the actual parsing
-and typing. We ensure that the typed AST we get is the same as the official
-OCaml compiler AST since we use the exact same frontend.
-
-% Toute une partie sur MLExplain lui-même...
-
-\subsection{Web interface changes}
-
-The data structures manipulated by MLExplain are very different from those
-manipulated by JSExplain since they don't interpret the same language. This led
-us to modify the web interface so that it can recognize the different syntactic
-elements of OCaml and pretty print them correctly as well as OCaml values.
-
-Fortunately the web interface is very simple. It is composed of some auxiliary
-JavaScript files, the main JavaScript file containing the logic of the
-application and the corresponding HTML page.
-
-\begin{figure*}[ht]
+\begin{figure*}[t]
   \centering
@@
 
+To interpret OCaml code, we need to parse and type it. However, our compiler
+does not allow us to use the standard library nor an external one. Fortunately,
+JSExplain itself did not trace both parsing and typing processes. That is why
+we have been able to separate totally the frontend from the interpreter itself
+and use another compiler to compile it. The figure \ref{fig:archi} describes
+the architecture of the whole application.
+
+The frontend is compiled with \emph{js\_of\_ocaml}
+\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
+\emph{Ocsigen} \cite{balat:hal-00691710}. The largest part of the frontend is a
+module that serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a
+JavaScript object compatible with the backend interpreter's own representation
+of the AST. We use the \texttt{compiler-libs} library to do the actual parsing
+and typing. We ensure that the typed AST we get is the same as the official
+OCaml compiler AST since we use the exact same frontend.
+
+\subsection{Solutions to compiler limitations}
+
+As stated before, the backend of our interpreter had to be written in
+\texttt{fOCaml} to be integrated into the web interface. This limitation forced
+to write non-idiomatic OCaml.
+
+% Toute une partie sur MLExplain lui-même...
+
+\subsection{Web interface changes}
+
+The data structures manipulated by MLExplain are very different from those
+manipulated by JSExplain since they don't interpret the same language. This led
+us to modify the web interface so that it can recognize the different syntactic
+elements of OCaml and pretty print them correctly as well as OCaml values.
+
+Fortunately the web interface is very simple. It is composed of some auxiliary
+JavaScript files, the main JavaScript file -- \texttt{navig-driver.js} --
+containing the logic of the application and the corresponding HTML page. The
+only modified file is \texttt{navig-driver.js}.
+
 \bibliographystyle{plain}

### 2018-05-24 paper.tex
done with JSExplain description

@@
   itself. This interpreter aims to provide the user with a better understanding
-  of the semantics of OCaml semantics.
+  of the semantics of OCaml.
 \end{abstract}
@@
 describe the execution of a JavaScript program by showing every step of an
-interpreter whose behavior is very closest to the specification of JavaScript.
-In this paper, we show how we adapted JSExplain to OCaml.
+interpreter whose behavior is very close to the specification of JavaScript. In
+this paper, we show how we adapted JSExplain to OCaml.
 
 Unlike JavaScript, OCaml has no official specification. However, OCaml code
-execution is fairly simple because the amount of language constructions is
+execution is fairly simple because the number of language constructions is
 small. We have written an interpreter for OCaml's typed abstract syntax tree
-(AST). This AST is close to the source code and need little transformations but
-supplies us with resolved names -- which is useful for module and signatures
-inclusion and absolutely mandatory for features like named parameters and
-optional parameters in functions. The semantic we give to OCaml is higher level
-than the one described in \emph{ZINC} \cite{Leroy-ZINC} virtual machine which is
-at the same level as bytecode.
+(AST). This AST is still close to the source code yet it provides additional
+crucial information. For instance, it mentions resolved names, which is useful
+for module and signatures inclusion and mandatory for interpreting features like
+named parameters and optional parameters in functions. The semantic we give to
+OCaml is higher level than the one described in \emph{ZINC} \cite{Leroy-ZINC}
+virtual machine, i.e., an interpreter for bytecode after compilation.
 
@@
 
-\subsection{Double Debugger}
+\subsection{A Double Debugger}
 \label{subsec:double-debugger}
@@
 
-ALAN: stopped here
+We instrument the compilation to JavaScript by adding code that logs events,
+namely entering a function, creating a scope, assigning a variable, and exiting
+a function. Each event captures the state, the stack, and the values of all
+local variables in scope of the interpreter code at the point where the event
+gets triggered. Code that is traced is compiled with this instrumentation,
+whereas code that is not traced, for instance supporting libraries, is directly
+compiled to JavaScript.
 
-JSExplain is a visual interpreter. To display its state, the interpreter is
-compiled with a compiler that adds calls to a tracing function. The traced
-retrieved during the execution of the program gives many information about the
-state of the interpreter:
+When using the tool, the source JavaScript program is parsed and run through the
+instrumented interpreter, thus generating a trace. We can then navigate this
+trace to see the state of the interpreter as well as the state of the
+interpreted program. Recovering the information about the interpreted code is
+not completely straightforward. For example, to recover the fragment of code to
+highlight, we find in the trace the closest previous event that contains a call
+to function with an argument named \texttt{\_term\_}. This argument corresponds
+to the AST of a subexpression, and this AST is decorated (by the parser) with
+locations. Note that, for efficiency reasons, we associate to each event from
+the trace its corresponding \texttt{\_term\_} argument during a single pass,
+performed immediately after the trace is produced.
 
-\begin{itemize}
-  \item existing variables and their values
-  \item return values of functions of the interpreter
-\end{itemize}
+Similarly, we are able to recover the state and environment associated with the
+event. The state of the interpreted program consists of four fields: the
+strictness flag, the value of the \texttt{this} keyword, the lexical
+environment, and the variable environment. We implemented a custom display for
+these elements, and also for values of the languages, in particular for objects:
+one may click on an object to reveal its contents and recursively explore it.
 
-JSExplain's web interface uses these traces in order to display the program and
-interpreter state at every step of the execution.
-
-The whole project takes the shape of a web application composed of the
-JavaScript interpreter and a view. The view allows the user to write JavaScript
-code and run it. It also displays the state of the program and the state of the
-interpreter.
-
-The program's execution is not interactive: when the button \texttt{run} is
-pressed, the program is executed all at one and the traces are generated. Only
-then, the user can replay the execution step by step.
-
-\subsection{JSExplain's compiler}
-
-% Référence pour coq ?
-JSExplain was first written in Coq. OCaml code has been extracted from the Coq
-source and is now the primary source code. The interpreter is compiled into
-JavaScript in order to use it within a web page. For this reason, the project
-embeds a homemade OCaml-to-JavaScript compiler.
-
-JSExplain also needs its own OCaml to JavaScript compiler in order to place the
-calls to \verb|log_event| which is the tracing function. This function is also
-used to get the location of the different syntactical objects in the source
-file in order to produce code highlighting in JSExplain's interface.
-
-This compiler offers several compilation modes:
-
-\begin{itemize}
-  \item \texttt{unlog}: generate JavaScript without tracing.
-  \item \texttt{log}: generate JavaScript with tracing.
-  \item \texttt{mlloc}, \texttt{token} and \texttt{ptoken}: generate files with
-    informations about syntastical elements from the input file.
-  \item \texttt{pseudo}: generate JavaScript enhanced with syntactic sugar for
-    pattern matching or monadic operations -- it is supposed to be easier to
-    read.
-\end{itemize}
-
-Modes \texttt{unlog} and \texttt{log} exist so that auxiliary code in the
-interpreter -- like linked-list or hash-map internal-management functions -- is
-not traced.  The user is thus able to concentrate on the logical part of the
-interpreter functioning and forget about implementation details.
-
-\subsubsection{fOCaml}
-This compiler does not recognized the whole OCaml language, it can only handle
-a small subset we named \texttt{fOCaml}. Among other restrictions, it cannot
-handle objects and imperative features, exceptions, inner modules, and tuples.
-The main reason for this limitation is that compiling OCaml into JavaScript is
-not trivial, many features -- e.g. nested pattern matching -- are hard to
-represent to JavaScript.
-
-The fOCaml code is compiled into \texttt{fJs}, a functional subset of
-JavaScript. The compiler preserves as much as possible the code structure to
-enable the user to understand both fOCaml and fJs codes as the same one.
-
-% Petite explication sur le fait que du coup on écrit du code fonctionnel pur
-% et les implications que ça a ?
+Finally, a web interface is presented to the user, to navigate in the trace.
+This navigation can be forward and backward, and at all times the source code
+and interpreter code are highlighted, and their states are displayed. We also
+provide a way to specify breakpoints, using arbitrary JavaScript expressions,
+that reference both the interpreter and interpreted programs. One may thus reach
+a point where some interpreter line is being evaluated with some source value
+satisfying a predicate, or vice-versa.
 

### 2018-05-24 paper.tex
tiny fixes
Continue la sous-section sur les limitations

@@
 At first glance, JSExplain seems tightly linked to JavaScript. However, the
-architecture of the project actually allows us to take off the interpreter
-completely and replace it with another one.
+architecture of the project actually lets us completely replace the interpreter
+with one for a different language.
 
@@
 
+% Difference entre les constructeurs, raison principale de la traduction à la
+% main
+
 \subsection{Solutions to compiler limitations}
@@
 
-% Toute une partie sur MLExplain lui-même...
+Error management could not be done using exceptions. This is why we used monadic
+computation. A monad is a data structure representing an action within a
+context. A monad typically take the form of a type constructor. It provides the
+following functions:
+
+\begin{verbatim}
+Let m, a monad:
+
+val return : 'a -> 'a m
+val bind : 'a m -> ('a -> 'b m) -> 'b m
+\end{verbatim}
+
+The function \texttt{return} let us build a new monadic value \verb|'a m| from
+a value of type \verb|'a|. The function \texttt{bind} (or operator \texttt{>>=}) is
+used to compose monads using a function of type \verb|'a -> 'b m|.
+
+This signature is typical from functions that may fail -- for example
+\verb|'a -> 'b option|. In this case, the \texttt{bind} function would apply
+the supplied function on the monad only if represents a valid state (see figure
+\ref{fig:error} for an implementation on \texttt{option}).
+
+\begin{figure}[b]
+  \begin{verbatim}
+let bind_option opt f = match opt with
+| Some value -> f value
+| None -> None
+  \end{verbatim}
+
+  \label{fig:error}
+  \caption{Implementation of operator bind on type option}
+\end{figure}
+
+Monads are an elegant way to manage errors, however it forces to nest many
+levels of lambda expressions to chain multiple calls to functions that may
+fail. This causes the source code to be extremely hard to read and therefore to
+maintain. Our compiler supports \texttt{ppx} extensions, hence we implemented a
+very simple syntax extension to transform a special let binding into a bind
+function call:
+
+\begin{verbatim}
+(* This beautiful syntax *)
+let%some a = function_that_could_fail 0 in
+continuation a
+
+(* becomes this *)
+bind_option
+  (function_that_could_fail 0)
+  (fun a -> continuation a)
+\end{verbatim}
 

### 2018-05-24 paper.tex
Ajouté un paragraphe sur les constructeurs de type

@@
 architecture of the project actually lets us completely replace the interpreter
-with one for a different language.
+with one for a different language. Our main task was to create a new OCaml
+interpreter that could be integrated into the rest of the application. Then, we
+noticed that the web application itself needed some changes -- for instance to
+pretty print values.
 
@@
 
-% Difference entre les constructeurs, raison principale de la traduction à la
-% main
+% TODO: Alan, si tu as une idée pour raccourcir ou découper cette phrase
+The main reason why both AST representations are not compatible and require an
+explicit hand-made serialization into JavaScript is how the compilers compile
+OCaml type constructors. Our compiler creates a JavaScript object from the type
+and treats constructors as methods building themselves the value. This value is
+a simple JavaScript object. \emph{js\_of\_ocaml}, on the other hand, uses its
+own internal representation of OCaml value that we cannot use directly.
 
@@
 computation. A monad is a data structure representing an action within a
-context. A monad typically take the form of a type constructor. It provides the
+context. A monad typically take the form of an algebraic type. It provides the
 following functions:
@@
 
-\begin{figure}[b]
+\begin{figure}[hb]
   \begin{verbatim}

### 2018-05-28 paper.tex
Continué de rédigé le papier
Ajout d'une fichier recensant les lignes de code, commentaires et appels à bind
Ajouté un tableaux de statistiques sur le projet
first draft

--- 2018-05-24 paper.tex
@@
 \usetikzlibrary{positioning,fit,shapes,arrows.meta}
+\usepackage{hyperref}
 
@@
 using a web-based
-tool.\footnote{https://jscert.github.io/jsexplain/branch/master/driver.html}
+tool.\footnote{\url{https://jscert.github.io/jsexplain/branch/master/driver.html}}
 
@@
 At first glance, JSExplain seems tightly linked to JavaScript. However, the
-architecture of the project actually allows us to take off the interpreter
-completely and replace it with another one.
-
-% Plus de blabla
+architecture of the project is quite modular, and can be used for a different
+programming language, by providing an interpreter written in the subset of OCaml
+that we support as well as a way to display OCaml values.
 
@@
 
-To interpret OCaml code, we need to parse and type it. However, our compiler
-does not allow us to use the standard library nor an external one. Fortunately,
-JSExplain itself did not trace both parsing and typing processes. That is why
-we have been able to separate totally the frontend from the interpreter itself
-and use another compiler to compile it. The figure \ref{fig:archi} describes
-the architecture of the whole application.
+To interpret OCaml code, we first need to parse and type it. However, our
+compiler does not allow us to use the standard library nor an external one. As
+we do not need to trace the parsing nor the typing, we have been able to
+separate totally the frontend from the interpreter itself and use another
+compiler to compile it. Figure \ref{fig:archi} describes the architecture of
+the whole application.
 
 The frontend is compiled with \emph{js\_of\_ocaml}
-\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project
-\emph{Ocsigen} \cite{balat:hal-00691710}. The largest part of the frontend is a
-module that serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a
-JavaScript object compatible with the backend interpreter's own representation
-of the AST. We use the \texttt{compiler-libs} library to do the actual parsing
-and typing. We ensure that the typed AST we get is the same as the official
-OCaml compiler AST since we use the exact same frontend.
+\cite{DBLP:journals/spe/VouillonB14}, the compiler of the project \emph{Ocsigen}
+\cite{balat:hal-00691710}. The largest part of the frontend is a module that
+serializes the \emph{Typedtree} -- the typed AST of OCaml -- into a JavaScript
+object compatible with the backend interpreter's own representation of the AST.
+We use the OCaml compiler, namely the \texttt{compiler-libs} library, to do the
+actual parsing and typing. We thus have the guarantee that the typed AST we get
+is the same as the official OCaml compiler AST.
 
-\subsection{Solutions to compiler limitations}
+Unfortunately, we cannot directly use the typed AST, as \emph{js\_of\_ocaml}
+takes OCaml bytecode as input, which does not include sufficient information
+such as the names of constructors. After parsing and typing, we thus navigate
+the AST and call JavaScript functions that create the data in the expected format.
 
-As stated before, the backend of our interpreter had to be written in
-\texttt{fOCaml} to be integrated into the web interface. This limitation forced
-to write non-idiomatic OCaml.
+\subsection{Compiler limitations}
 
-% Toute une partie sur MLExplain lui-même...
+As stated before, the backend of our interpreter is written in a purely
+functional strict subset of OCaml. As a consequence, it was necessary to encode
+some features, such as exceptions. To this end, we use a monadic approach and
+rely on \texttt{ppx} extensions to simplify the syntax.
+
+As an example, consider the \texttt{option} monad, whose main operator is
+defined as follows.
+
+\begin{verbatim}
+let bind_option opt f = match opt with
+| Some value -> f value
+| None -> None
+\end{verbatim}
+
+To use this operator, one would need to write the following.
+\begin{verbatim}
+bind_option
+  (function_that_could_fail 0)
+  (fun a -> continuation a)
+\end{verbatim}
+Such code can become difficult to read, especially when one is chaining several
+functions.
+
+The use of \texttt{ppx} lets us write the much simpler code, which is
+transformed to the previous version during parsing.
+\begin{verbatim}
+let%some a = function_that_could_fail 0 in
+continuation a
+\end{verbatim}
 
@@
 The data structures manipulated by MLExplain are very different from those
-manipulated by JSExplain since they don't interpret the same language. This led
-us to modify the web interface so that it can recognize the different syntactic
+manipulated by JSExplain since they don't interpret the same language. We thus
+had to modify the web interface so that it can recognize the different syntactic
 elements of OCaml and pretty print them correctly as well as OCaml values.
@@
 containing the logic of the application and the corresponding HTML page. The
-only modified file is \texttt{navig-driver.js}.
+only modified file is \texttt{navig-driver.js}. The resulting tool can be tested
+online.\footnote{\url{https://docteur-lalla.github.io/mlexplain/branch/master/driver.html}}
+
+\section{Statistics}
+
+We give some statistics about the project in Figure~\ref{fig:statistics}. As can
+be seen, the largest part of the code is the frontend, in particular the
+conversion from OCaml's AST to our format. The interpreter is quite short,
+because the execution of OCaml is actually very simple. Note that we support all
+of OCaml, with the exception of objects.
+
+\begin{figure}[hb]
+  \begin{tabular}{| c | c | c | c |}
+    \hline
+    & Comments & Code & Bind calls \\
+    \hline
+    Interpreter & 164 & 394 & 62 \\
+    Primitives & 31 & 35 & 0 \\
+    Execution context & 22 & 13 & 0 \\
+    Frontend & 41 & 542 & 0 \\
+    Syntax definition & 17 & 65 & 0 \\
+    Auxiliary files & 72 & 232 & 2 \\
+    \hline
+    Total (no .mli) & 392 & 1621 & 70 \\
+    Total (with .mli) & 401 & 1636 & 70 \\
+    \hline
+  \end{tabular}
+
+  \centering
+  \caption{Statistics about the project's code}
+  \label{fig:statistics}
+\end{figure}
+
+\section{Conclusion}
+We have shown how JSExplain can be easily adapted to other programming language,
+in this case OCaml. Most of the complexity of OCaml being its typing, writing
+the interpreter was quite straightforward. As future work, we plan to extend the
+subset of OCaml supported by our compiler, to be able to directly use the
+standard library when writing interpreters.
 

### 2018-05-28 paper.tex
Ajoute un bout à la conclusion

@@
 in this case OCaml. Most of the complexity of OCaml being its typing, writing
-the interpreter was quite straightforward. As future work, we plan to extend the
-subset of OCaml supported by our compiler, to be able to directly use the
-standard library when writing interpreters.
+the interpreter was quite straightforward. The current implementation of our
+interpreter is very simple and contains a lot of documentation. The
+architecture of the code is also straightforward, which allows anyone to easily
+extend the interpreter. The simple design of our interpreter makes it a perfect
+candidate to base a specification of OCaml on. MLExplain also provides a
+confortable environment to experiment with the semantics of OCaml. It is thus a
+perfect playground for future evolution of the language.
+
+As future work, we plan to extend the subset of OCaml supported by our
+compiler, to be able to directly use the standard library when writing
+interpreters.
 

### 2018-05-28 paper.tex
talk about record patterns
final tweaks?

--- 2018-05-28 paper.tex
@@
 At first glance, JSExplain seems tightly linked to JavaScript. However, the
-architecture of the project is quite modular, and can be used for a different
-programming language, by providing an interpreter written in the subset of OCaml
-that we support as well as a way to display OCaml values.
+architecture of the project is quite modular and can be used for a different
+programming language, by providing both an interpreter written in the subset of
+OCaml that we support and a way to display OCaml values.
 
@@
 To interpret OCaml code, we first need to parse and type it. However, our
-compiler does not allow us to use the standard library nor an external one. As
-we do not need to trace the parsing nor the typing, we have been able to
-separate totally the frontend from the interpreter itself and use another
-compiler to compile it. Figure \ref{fig:archi} describes the architecture of
-the whole application.
+compiler does not allow us to use the standard library nor an external one, as
+they are not written in the subset of OCaml that we support. As we do not need
+to trace the parsing nor the typing, we have been able to separate totally the
+frontend from the interpreter itself and use another compiler to compile it.
+Figure \ref{fig:archi} describes the architecture of the whole application.
 
@@
 
-The use of \texttt{ppx} lets us write the much simpler code, which is
-transformed to the previous version during parsing.
+The use of \texttt{ppx} lets us write much simpler code, which is transformed to
+the previous version during parsing.
 \begin{verbatim}
@@
 because the execution of OCaml is actually very simple. Note that we support all
-of OCaml, with the exception of objects.
+of OCaml, with the exception of objects and record patterns.
 
@@
 in this case OCaml. Most of the complexity of OCaml being its typing, writing
-the interpreter was quite straightforward. As future work, we plan to extend the
-subset of OCaml supported by our compiler, to be able to directly use the
-standard library when writing interpreters.
+the interpreter was quite straightforward. The current implementation of our
+interpreter is very simple and contains a lot of documentation. The architecture
+of the code is as unsurprising as possible, enabling easy extensions. As the
+design of our interpreter is very simple, it could be considered as a first step
+towards a specification of OCaml. MLExplain also provides a comfortable
+environment to experiment with the semantics of OCaml. It is thus a good
+playground for future evolution of the language.
+
+As future work, we plan to extend the subset of OCaml supported by our compiler,
+to be able to directly use the standard library when writing interpreters, and
+to implement the few OCaml construct missing from our interpreter. We also want
+to improve the web interface and to allow the user to load code from multiple files.
 

### 2018-07-23 paper.tex
added the reviews, started to address them
adding screenshot, minted

@@
 %\documentclass{article}
-\documentclass[twocolumn]{article}
+\documentclass[twocolumn,a4paper]{article}
 \usepackage[utf8]{inputenc}
+\usepackage[top=1cm, bottom=2cm, left=1.5cm, right=1.5cm]{geometry} 
+\usepackage{minted}
 \usepackage{tikz}
@@
 
-\author{ K\'evin Le Bon \and Alan Schmitt }
-
+\author{ K\'evin Le Bon (Inria) \and Alan Schmitt (Inria) }
 \title{MLExplain}
+\date{}
 
@@
 
+Note that JSExplain, and by extension MLExplain, runs a program to its
+completion to generate the trace, and it is that trace that is explored using
+the web interface. It thus does not currently offer the ability to alter the
+state of the program as it is being explored.
+
 \section{MLExplain}
@@
 
-\begin{figure*}[t]
+\begin{figure*}
   \centering
@@
 
-\begin{verbatim}
+\begin{minted}{ocaml}
 let bind_option opt f = match opt with
@@
 | None -> None
-\end{verbatim}
+\end{minted}
 
 To use this operator, one would need to write the following.
-\begin{verbatim}
+\begin{minted}{ocaml}
 bind_option
@@
   (fun a -> continuation a)
-\end{verbatim}
+\end{minted}
 Such code can become difficult to read, especially when one is chaining several
@@
 the previous version during parsing.
-\begin{verbatim}
+\begin{minted}{ocaml}
 let%some a = function_that_could_fail 0 in
 continuation a
-\end{verbatim}
+\end{minted}
 
@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=14cm]{mlexplain.png}
+  \caption{MLExplain}
+  \label{fig:mlexplain}
+\end{figure*}
+
 \section{Statistics}
@@
 
-\begin{figure}[hb]
+\begin{figure}
   \begin{tabular}{| c | c | c | c |}
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% TeX-engine: latex_sh
+%%% End:

### 2018-07-23 paper.tex
addressing more things

@@
 
-\section{MLExplain}
-
-At first glance, JSExplain seems tightly linked to JavaScript. However, the
-architecture of the project is quite modular and can be used for a different
-programming language, by providing both an interpreter written in the subset of
-OCaml that we support and a way to display OCaml values.
-
-\subsection{Two compilers}
-
 \begin{figure*}
@@
 
-  \caption{Architecture of MLExplain}
-  \label{fig:archi}
+  \caption{Architecture of MLExplain}\label{fig:archi}
 \end{figure*}
+
+\section{MLExplain}
+
+At first glance, JSExplain seems tightly linked to JavaScript. However, the
+architecture of the project is quite modular and can be used for a different
+programming language, by providing both an interpreter written in the subset of
+OCaml that we support and a way to display OCaml values.
+
+\subsection{Two compilers}
 
@@
 only modified file is \texttt{navig-driver.js}. The resulting tool can be tested
-online.\footnote{\url{https://docteur-lalla.github.io/mlexplain/branch/master/driver.html}}
+online\footnote{\url{https://docteur-lalla.github.io/mlexplain/branch/master/driver.html}}
+and is shown in Figure~\ref{fig:mlexplain}. The top-left panel shows the
+interpreted, here a simple eval function using GADTs, with the state of the
+program depicted on the right. The bottom panel shows the
+interpreter, with its state below. One can see that at this point the
+\texttt{eval} identifier is being evaluated, with the values of \texttt{r},
+\texttt{l}, and \texttt{b} shown on the right.
 
@@
 \end{figure*}
+
+\subsection{Design Choices}\label{subsec:design}
+
+We now motivate our design choices. First, we use JavaScript because MLExplain
+is a witness that JSExplain can easily be adapted to other languages, by
+providing three components: an interpreter written in the subset of OCaml that
+our compiler supports, a parser of the language, and a library to display OCaml
+values. It would be possible to sever all links to the infrastructure provided
+by JSExplain, but this would require significant work.
+
+It is possible to use an existing debugger to explore how an interpreter works.
+This is how we initially debugged our JavaScript interpreter and this was the
+motivation to start working on JSExplain. Indeed, our interpreter is written as
+a purely functional set of recursive functions that extensively use monadic
+operations implemented in continuation passing style. It is difficult to
+correctly navigate such functions, as stepping over a continuation means
+reaching the end of the program. One thus has to be aware of whether a function
+call is a continuation or a usual function. In addition, we did not find an easy
+way to display the state of the interpreted program, even when relying on the
+use of fixed names for the term, the state, and the context.
 

### 2018-07-23 paper.tex
I think I have addressed everything

@@
 \usepackage[utf8]{inputenc}
-\usepackage[top=1cm, bottom=2cm, left=1.5cm, right=1.5cm]{geometry} 
+\usepackage[top=1.5cm, bottom=2cm, left=2cm, right=2cm]{geometry} 
 \usepackage{minted}
@@
 providing three components: an interpreter written in the subset of OCaml that
-our compiler supports, a parser of the language, and a library to display OCaml
-values. It would be possible to sever all links to the infrastructure provided
-by JSExplain, but this would require significant work.
+our compiler supports, a parser of the language, and a library to display
+runtime values. It would be possible to sever all links to the infrastructure
+provided by JSExplain, but this would require significant work.
 
-It is possible to use an existing debugger to explore how an interpreter works.
-This is how we initially debugged our JavaScript interpreter and this was the
-motivation to start working on JSExplain. Indeed, our interpreter is written as
-a purely functional set of recursive functions that extensively use monadic
+Second, it is possible to use an existing debugger to explore how an interpreter
+works. This is how we initially debugged our JavaScript interpreter and this was
+the motivation to start working on JSExplain. Indeed, our interpreter is written
+as a purely functional set of recursive functions that extensively use monadic
 operations implemented in continuation passing style. It is difficult to
@@
 reaching the end of the program. One thus has to be aware of whether a function
-call is a continuation or a usual function. In addition, we did not find an easy
-way to display the state of the interpreted program, even when relying on the
-use of fixed names for the term, the state, and the context.
+call involves a continuation or a usual function. In addition, we did not find
+an easy way to display the state of the interpreted program, even when relying
+on the use of fixed names for the term, the state, and the context. The use of
+fixed names may seem fragile, but as we only rely on a few of them and as the
+signature of the evaluation functions are uniform, we have not observed any
+issue with this approach.
 
@@
 As future work, we plan to extend the subset of OCaml supported by our compiler,
-to be able to directly use the standard library when writing interpreters, and
-to implement the few OCaml construct missing from our interpreter. We also want
-to improve the web interface and to allow the user to load code from multiple files.
+to be able to directly use the standard library when writing interpreters. We
+also plan to implement the few OCaml construct missing from our interpreter for
+lack of time, such as record patterns and objects. It would be most useful to
+document the evaluation order of every OCaml construction to make sure our
+interpreters follows it. Finally, be want to improve the web interface and to
+allow the user to load code from multiple files.
 

### 2018-07-23 paper.tex
final tweaks

@@
 and is shown in Figure~\ref{fig:mlexplain}. The top-left panel shows the
-interpreted, here a simple eval function using GADTs, with the state of the
+interpreted program, here a simple eval function using GADTs, with the state of the
 program depicted on the right. The bottom panel shows the
@@
   \centering
-  \includegraphics[width=14cm]{mlexplain.png}
+  \includegraphics[width=17cm]{mlexplain.png}
   \caption{MLExplain}
@@
 because the execution of OCaml is actually very simple. Note that we support all
-of OCaml, with the exception of objects and record patterns.
+of OCaml, with the exception of objects and record patterns, which we have not
+had the time to implement.
 
@@
 to be able to directly use the standard library when writing interpreters. We
-also plan to implement the few OCaml construct missing from our interpreter for
-lack of time, such as record patterns and objects. It would be most useful to
-document the evaluation order of every OCaml construction to make sure our
-interpreters follows it. Finally, be want to improve the web interface and to
-allow the user to load code from multiple files.
+also plan to implement the few OCaml constructs missing from our interpreter. In
+addition, it would be most useful to document the evaluation order of every
+OCaml construction to make sure our interpreters follows it. Finally, we want to
+improve the web interface and to allow the user to load code from multiple
+files.
 

### 2018-07-25 paper.tex
Add reference to JSExplain in the abstract

@@
   user to inspect both their program's state and the interpreter's state
-  itself. This interpreter aims to provide the user with a better understanding
-  of the semantics of OCaml.
+  itself. This interpreter is derived from JExplain, a step-by-step interpreter
+  for JavaScript. Its original goal is to prove that JSExplain can easily be
+  reused with another language. MLExplain also aims to provide the user with a
+  better understanding of the semantics of OCaml.
 \end{abstract}

### 2018-09-28 paper.tex
old changes I forgot to commit

@@
-%\documentclass{article}
-\documentclass[twocolumn,a4paper]{article}
-\usepackage[utf8]{inputenc}
+% \documentclass{article}
+\documentclass[twocolumn,a4paper]{article} \usepackage[utf8]{inputenc}
 \usepackage[top=1.5cm, bottom=2cm, left=2cm, right=2cm]{geometry}
-\usepackage{minted}
-\usepackage{tikz}
-\usetikzlibrary{positioning,fit,shapes,arrows.meta}
-\usepackage{hyperref}
+\usepackage{minted} \usepackage{tikz}
+\usetikzlibrary{positioning,fit,shapes,arrows.meta} \usepackage{hyperref}
 
-\author{ K\'evin Le Bon (Inria) \and Alan Schmitt (Inria) }
-\title{MLExplain}
+\author{ K\'evin Le Bon (Inria) \and Alan Schmitt (Inria) } \title{MLExplain}
 \date{}
@@
 \begin{abstract}
-  MLExplain is a step-by-step interpreter for OCaml that enables the
-  user to inspect both their program's state and the interpreter's state
-  itself. This interpreter is derived from JExplain, a step-by-step interpreter
-  for JavaScript. Its original goal is to prove that JSExplain can easily be
-  reused with another language. MLExplain also aims to provide the user with a
-  better understanding of the semantics of OCaml.
+  MLExplain is a step-by-step interpreter for OCaml that enables the user to
+  inspect both their program's state and the interpreter's state itself. This
+  interpreter is derived from JExplain, a step-by-step interpreter for
+  JavaScript. The original goal of this work is to show that JSExplain can
+  easily be reused with another language. MLExplain also aims to provide the
+  user with a better understanding of the semantics of OCaml.
 \end{abstract}
